---
title: "Simulating binary data"
output:
  html_document:
    css: webex.css
    includes:
      after_body: webex.js
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("webex")) {
  stop("You must have the 'webex' package installed to knit HTML from this template.\n   devtools::install_github(\"dalejbarr/webex\")")
} else {
  library("webex")
}
```

If you know how to simulate data for a study with a continuous DV, you are just one step away from simulating data with a binary DV.

1. Create your code to simulate continuous data, except don't add in the final Gaussian residuals.
2. Treat the model fits as values in log odds space, and write a function that iterates over the model predictions, converts them to probabilities, and samples 1, 0 based on the probability.

## Create the data set

So, here's an example.  Let's say that you have created the following simulated data set:

```{r sim-data1, message=FALSE}
library("tidyverse")
set.seed(999)

mu <- 1.1

dat <- tibble(subject_id = rep(1:10, each = 2),
              condition = rep(c("control", "experimental"), 10),
              sri = rep(rnorm(10), each = 2), # by-subject random intercept
              eff = rep(c(-.11, .11), times = 10), # fixed effect of condition
              eta = mu + sri + eff) # the predicted value (no residual)

dat
```

Now, let's define a function that (1) converts log odds (`eta`) to probabilities and then (2) samples 0s, 1s based on those probability.

## Create a function to generate Bernoulli trials

```{r sample}
bernoulli_trial <- function(eta) {
  p <- (1 / (1 + exp(-eta))) # inverse logit function
  sample(c(1, 0), 1, prob = c(p, 1 - p)) # choose 1 or 0 based on p
}

# try it out
replicate(1000, bernoulli_trial(-1)) %>% table()
replicate(1000, bernoulli_trial(0)) %>% table()
replicate(1000, bernoulli_trial(1)) %>% table()
```

## Add the column to your data

Now we will make a new column by running this function using `purrr::map_dbl()` from tidyverse. (The `map_*()` functions are iterating functions. See [the `purrr` documentation](https://purrr.tidyverse.org/) for more info.) 

The binary response variable is stored in `Y`.

```{r final}
dat %>%
  mutate(Y = map_dbl(eta, bernoulli_trial))
```

