---
title: "Contingency Tables and Chi-Square Test"
author: "Dale Barr"
date: "4/8/2021"
output: webex::webex_default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Contingency tables and chi-square

### Contingency tables

You can do this in base R using `xtabs()` (for crosstabs) or by using the tidyverse, with `dplyr::count()`, optionally followed by reshaping with `tidyr::spread()`.

#### Contingency table in base R with `xtabs()`:

```{r xtabs}
library("tidyverse")

xtabs(~ eye_color + gender, starwars)
```

- pros: easy
- cons: results table not guaranteed to be tidy

#### Contingency table with tidyverse
  
- pros: more flexible, and results in a tidy tibble
- cons: more typing

**NB: Major change in behavior with factors in dplyr version 0.8.0 (released in 2019)**.

  - If you use `count()` with the defaults, then it will count only those levels appearing in the data. In other words, counts of zero are not possible.

  - If you use `count()` with the argument `.drop = FALSE`, *and your variables are defined as factors* then it will count all levels, including combinations that don't appear in the data (they will appear in the output with `n = 0`).

See the section entitled [Give factors some love](https://www.tidyverse.org/articles/2019/02/dplyr-0-8-0/) in this article on dplyr.

How do you know which version of dplyr you have? Here's a quick way to check.

```{r pkg-vers}
packageVersion("dplyr")
```

If your version is less than 0.8.0 then this doesn't apply (and you should upgrade; use `tidyverse::tidyverse_update()`).

Here is the behavior of `count()` with the defaults.

```{r tidy-xtabs-dplyr-drop}
starwars %>%
  filter(!is.na(gender)) %>%
  count(gender, eye_color) 
```

Now compare this output to:

```{r tidy-xtabs-dplyr-nodrop}
## NB: this will only work with dplyr >= 0.8.0
starwars %>%
  filter(!is.na(gender)) %>%
  mutate(gender = factor(gender), eye_color = factor(eye_color)) %>% # make factors
  count(gender, eye_color, .drop = FALSE) 
```

If you want to make it into a proper table, you can reshape it using `tidyr::spread()`:

```{r sw-spread}
sw_data <- starwars %>%
  filter(!is.na(gender)) %>%
  mutate(gender = factor(gender), eye_color = factor(eye_color)) %>% # make factors
  count(gender, eye_color, .drop = FALSE) 

spread(sw_data, key = gender, value = n)
```

### Chi-square test for independence

Use `chisq.test()`. Note that counts in the table `x` must be independent. `x` can be a vector or a matrix/table.

```{r chi-sq-test}
  ## example from chisq.test documentation
  ## From Agresti(2007) p.39
  M <- as.table(rbind(c(762, 327, 468), c(484, 239, 477)))
  dimnames(M) <- list(gender = c("F", "M"),
		      party = c("Democrat","Independent", "Republican"))
  (Xsq <- chisq.test(M))  # Prints test summary
  Xsq$observed   # observed counts (same as M)
  Xsq$expected   # expected counts under the null
```



[^1]: Furnham, A. (1986), The Robustness of the Recency Effect: Studies Using Legal Evidence. *The Journal of General Psychology*, *113*, 351--357.

