---
title: "One-way ANOVA with regression"
output: html_document
---

You conduct a study looking at the effect of drinking different beverages on response times in a simple task where participants are instructed to press a button in response to a flashing light. You measure the mean response time for each subject, after drinking one of three drinks: (1) water (control condition); (2) coffee; (3) wine. The factor `group` is a between-subjects variable. The data are available in the file [beverages.csv](beverages.csv).

## Task 1

Import the data and create boxplots showing summary information for mean RT in each of the three groups.

```{r import-and-plot}
```

## Task 2

Use the `aov()` function to run a one-way ANOVA on the data.

```{r one-way-aov}
```

## Task 3

Now repeat the analysis, except using regression (R function `lm()`) instead of `aov()`. NOTE: you should create your own dummy variable(s) to code the `group` variable. HINT: you wil need to use model comparison to test the main effect.

```{r one-way-lm}
```

If you have done this correctly, you should get the same $p$-value for both analyses.
