---
title: "Formative Assignment 5: Simulating Linear Mixed-Effects Data"
author: "L3 Teaching Team"
output: html_document
---

```{r setup, include=FALSE}
# DO NOT REMOVE THIS BLOCK NOR ALTER ANYTHING IN THIS BLOCK!
knitr::opts_chunk$set(echo = TRUE)

# these functions are here for validation
.chk_numeric <- function(x) {if (exists(x)) {is.numeric(get(x))} else FALSE}
.chk_tibble <- function(x) {if (exists(x)) {is.tibble(get(x))} else FALSE}
.chk_matrix <- function(x) {if (exists(x)) {is.matrix(get(x))} else FALSE}
.has_only_cols <- function(x, y) {
  if (.chk_tibble(x)) {
    setequal(names(get(x)), y)
  } else FALSE
}

.my_seed <- 83154549L
set.seed(.my_seed)
.assign_id <- 
"06_lmem1"

library("tidyverse") # you're welcome
```

# Part I: Interpreting output

Consider the output below from a linear mixed-effects analysis of data simulated based on the `lme4::sleepstudy` data.

## GLM

$$Y_{ij} = \beta_0 + \beta_1 X_{ij} + e_{ij}$$

$$\beta_0 = \gamma_{00} + S_{0i}$$

$$\beta_1 = \gamma_{10} + S_{1i}$$

$$\left< S_{0i}, S_{1i} \right> \sim N\left(\left<0, 0\right>, \mathbf \Sigma\right)$$

$$\mathbf\Sigma = \left(\begin{array}{cc}{\tau_{00}}^2 & \rho\tau_{00}\tau_{11}\\\rho\tau_{00}\tau_{11} & {\tau_{11}}^2\end{array}\right)$$

$$e_{ij} \sim N(0, \sigma^2)$$


## Output

```
Linear mixed model fit by REML ['lmerMod']
Formula: Reaction ~ Day + (Day | Subject)
   Data: dat

REML criterion at convergence: 1822.1

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.33628 -0.52823 -0.05326  0.60517  2.13807 

Random effects:
 Groups   Name        Variance Std.Dev. Corr 
 Subject  (Intercept) 1077.2   32.82         
          Day         1225.2   35.00    -0.06
 Residual              718.9   26.81         
Number of obs: 180, groups:  Subject, 18

Fixed effects:
            Estimate Std. Error t value
(Intercept)  -32.980      8.581  -3.843
Day            9.341      8.280   1.128

Correlation of Fixed Effects:
    (Intr)
Day -0.088
convergence code: 0
Model failed to converge with max|grad| = 0.00372352 (tol = 0.002, component 1)

```

Answer the following questions about the output.

The parameter estimate for $\tau_{00}$ equals:

```{r output_01}
t00 <- NULL
```

The parameter estimate for ${\tau_{11}}^2$ equals:

```{r output_02}
t11_sq <- NULL
```

The grand intercept $\gamma_{00}$ is:

```{r output_03}
gamma_00 <- NULL
```

The grand slope $\gamma_{11}$ is:

```{r output_04}
gamma_10 <- NULL
```

The error variance $\sigma^2$ is:

```{r output_05}
sig_sq <- NULL
```

# Part II: Data simulation

In this part, we will be making our own simulated `sleepstudy` data for 18 subjects.  Your parameter values should reflect the data-generating process below (knit to HTML to better see):

$$Y_{ij} = \beta_0 + \beta_1 X_{ij} + e_{ij}$$
$$\beta_0 = 310 + S_{0i}$$
$$\beta_1 = 4 + S_{1i}$$
$$\left< S_{0i}, S_{1i} \right> \sim N\left(\left<0, 0\right>, \mathbf \Sigma\right)$$
$$\mathbf\Sigma = \left(\begin{array}{cc}{30}^2
& (0.9)(30)(33)\\
(0.9)(30)(33) & 
{33}^2\end{array}\right)$$
$$e_{ij} \sim N(0, 25^2)\\$$
<!-- assessr: datasim -->

We will proceed in three steps.

1. Generate random effects (intercept and slope offsets) for 18 subjects from a bivariate normal distribution using the parameter values for the variance components.

2. Generate a tibble containing the fixed effects $\gamma_{00}$ and $\gamma_{10}$, and random noise according to the value of $\sigma^2$.

3. Combine the random effects generated in part 1 with the fixed effects and error generated in part 2, and calculate the $Y$ values according to the DGP.

## Step 1: Generate random effects

You will need to generate data for the matrix `rmx`, where the first column should have the random intercepts $S_{0i}$ and the second column should have the random slopes $S_{1i}$.

```{r make_vcov}
## TODO: write code to randomly generate 18 bivariate observations

## Hint: How do you randomly generate bivariate data based on a covariance matrix?
## You've done this before, most recently on PST1.

## 'rmx' should be an 18x2 matrix
## the call to matrix() is just a placeholder
## it shows you how the resulting matrix should be structured
## you should be using a different function to create the matrix

rmx <- matrix(rep(0, 36), nrow = 18,
              dimnames = list(NULL, c("S_0s", "S_1s")))
```

## Step 2: Make a tibble with fixed effects and random error (noise)

In the next chunk, make a table containing the values for $\gamma_{00}$, $\gamma_{10}$, and $e_{ij}$.

This should be a tibble with 180 observations (18 subjects, observed over 10 days).

```{r obs}
obs <- tibble(Subject = factor(rep(seq_len(18), each = 10)),
              Days = rep(0:9, 18),
              gamma_00 = 0,  # replace with correct values
              gamma_10 = 0,  # replace with correct values
              e_ij = 0)      # replace with correct values
```

## Step 3: Combine

Now find a way to combine the information in the `obs` table with the info in `rmx` so that you create a table that looks like the `sleepstudy` table. (*Hint: you know how to columns in two tables already (two-table tidyverse verb), but `rmx` is a matrix rather than a tibble. It needs to be converted to a tibble before you combine it with the other table. You can convert it using `as_tibble()`.*)

The resulting `simdata` table should *only* have the columns `Reaction`, `Days`, and `Subject`. Hint: check the simulated tables of data in the [multilevel web app](http://shiny.psy.gla.ac.uk/Dale/multilevel).

```{r combine}
simdata <- NULL

# print it out
simdata
```

# Validation

The rest of this file simply validates the answers you entered above, and checks that you are working with the right data.

Knit your file and look at the resulting HTML. If you pass all of the checks below, you are ready to submit!

- `t00` exists and is numeric: `r .chk_numeric("t00")`
- `t11_sq` exists and is numeric: `r .chk_numeric("t11_sq")`
- `gamma_00` exists and is numeric: `r .chk_numeric("gamma_00")`
- `gamma_11` exists and is numeric: `r .chk_numeric("gamma_10")`
- `sig_sq` exists and is numeric: `r .chk_numeric("sig_sq")`
- `rmx` exists and is a numeric matrix: `r .chk_matrix("rmx")`
- `obs` exists and is a tibble: `r .chk_tibble("obs")`
- `simdata` exists and is a tibble: `r .chk_tibble("simdata")`
- `simdata` has only the columns `Reaction`, `Days`, and `Subject`: `r .has_only_cols("simdata", c("Subject", "Days", "Reaction"))`
