---
title: "Miscellaneous extra things"
author: "Dale"
date: "4 April 2019"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Window functions in dplyr

```{r window-fns, message=FALSE}
library("tidyverse")
library("babynames")

babynames %>%
  group_by(year, sex) %>%
  filter(near(prop, max(prop))) %>%
  ungroup()
```


## Making tibbles by hand

```{r tibs}
# tibble way
tibble(subject = 1:8,
       A = rep(c("A1", "A2"), each = 4),
       Y = rnorm(8))

# tribble way
tribble(~subject, ~A, ~Y,
        1, "A1", rnorm(1),
        2, "A2", rnorm(1))
```


## Inline R code in RMarkdown

```{r t-test}
set.seed(999)
v1 <- rnorm(100)
v2 <- rnorm(100)

result <- t.test(v1, v2, var.equal = TRUE)

x <- broom::tidy(result)
x
```

**Source**: The result of the t-test was: 
t(`` `r knitr::inline_expr("x$parameter")` ``) = 
`` `r knitr::inline_expr("round(x$statistic, 3)")` ``, 
p = `` `r knitr::inline_expr("round(x$p.value, 3)")` ``.

**Generated**: The result of the t-test was: 
t(`r x$parameter`) = `r round(x$statistic, 3)`, p = `r round(x$p.value, 3)`.


## Recoding and re-leveling

- syntax: `recode(varname, current = "New", ...)`

```{r recode1}
vv <- sample(rep(c("C 1", "E"), each = 5))
dplyr::recode(vv, `C 1` = "Control", E = "Experimental")
```

## Releveling factors

- syntax: `forcats::fct_relevel(varname, "First Level", "Second Level")`

```{r fct-relevel}
vvv <- rep(c("Antidepressant", "Placebo", "Other"), each = 3)
fct_relevel(vvv, "Placebo")
```

## Some useful additional functions

```{r misc}
## names(): get the column names
names(starwars)

## str(): get the structure of the object
# str(starwars)

## class(): what type of object is it?
class(starwars)

## nrow(): how many rows in the table?
nrow(starwars)

## ncol(): how many columns?
ncol(starwars)

## dim(): what are the table dimensions? 
dim(starwars)

## summary(): get summaries of all columns
summary(iris)

## distinct(): get distinct combinations in the data
distinct(mtcars, am, gear, carb)

## count(): count observations for each distinct combo
count(mtcars, am, gear, carb)

## select helpers: starts_with()
select(who, starts_with("new_sp"))

## test values
## is.logical(); is.numeric(); is.character(); is.factor()
## as.integer(); as.numeric(); as.character(); as.factor()
is.logical(starwars$name)
is.character(starwars$name)
is.factor(starwars$name)

## levels()
x <- factor(starwars$name)
levels(x)

## summarizing functions; often used with summarize
## min(); max(); sum(); dplyr::n(); median(); rank()
mtcars %>%
  summarize(mn = min(mpg), mx = max(mpg), med = median(mpg))
```

```{r more, eval=FALSE}
## viewing tables
## View(); glimpse()
## head(); tail()
print(starwars) # same as just typing "starwars"
print(mtcars, n = +Inf) # print all rows! see ?print.tbl
```

```{r even-more}
## repeating things and making sequences
## rep(); seq()
rep(letters[1:5], 5)
rep(LETTERS[1:5], each = 5)

seq(1, 10, .5)
seq(10, 1, length.out = 7)
```

