---
title: "Data Wrangling Exercise: Scoring the Autism Quotient (AQ)"
output:
  html_document:
    css: webex.css
    includes:
      after_body: webex.js
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("webex")) {
  stop("You must have the 'webex' package installed to knit HTML from this template.\n   devtools::install_github(\"dalejbarr/webex\")")
} else {
  library("webex")
}
```

# Background: Data Wrangling

Here’s a chance to test your data analysis skills in R.

For a research project, we had prospective participants complete the short 10-item version of the Autism-Spectrum Quotient (AQ) (Baron-Cohen, Wheelwright, Skinner, Martin, & Clubley, 2001), which is designed to measure autistic traits in adults. We used Survey Monkey to host the survey. The data from Survey Monkey was downloaded as a .csv file, and has been anonymized.

|     |                                               |
|-----|-----------------------------------------------|
| Q01	|I often notice small sounds when others do not.|
| Q02	|I usually concentrate more on the whole picture, rather than small details.|
| Q03	|I find it easy to do more than one thing at once.|
| Q04	|If there is an interruption, I can switch back to what I was doing very quickly.|
| Q05	|I find it easy to read between the lines when someone is talking to me.|
| Q06	|I know how to tell if someone listening to me is getting bored.|
| Q07	|When I'm reading a story, I find it difficult to work out the characters' intentions.|
| Q08	|I like to collect information about categories of things.|
| Q09	|I find it easy to work out what someone is thinking or feeling just by looking at their face.|
| Q10	|I find it difficult to work out people's intentions.|


Responses to each item were measured on a four-point scale: Definitely Disagree, Slightly Disagree, Slightly Agree, Definitely Agree. To avoid response bias, each question is scored according to one of two different formats. For questions 1, 7, 8, and 10, a point is assigned for agreement (either “Slightly Agree” or “Definitely Agree”), zero otherwise. For the remaining questions, a point is assigned for disagreement (“Slightly Disagree” or “Definitely Disagree”), zero otherwise. The AQ for each participant is just the total points across all 10 questions.

Your task is to develop an R script to calculate the AQ score for the (anonymized) responses in this data file. The script should be written in a general manner so that if you were to add more participants through Survey Monkey, re-download the csv file, and run the script on it, it would run without error.

After calculating the AQ scores for each participant, plot a distribution of the scores using ggplot2.

## Steps to the solution

1. Open a new RMarkdown file and create a new code chunk (Ctrl-Alt-I, or `Code` | `Insert Chunk` from the pulldown menu). In this first chunk, prepare your session by loading in any add-on packages you need. Then Load the data from the CSV file into a variable `dat` in your session.

`r hide("Solution")`
  
```{r}
  library("tidyverse")
  
  dat <- read_csv("name_of_my_file.csv")
```
  
s`r unhide()`

2. In the console, *look at your data*.  It is important to get a sense of how the data are structured before going further.

  `r hide("Hint")`
  There are several ways to look at your data. The one I prefer is just to print it in the console by typing the variable name `dat`.
  
  If the table is too wide, you can get a rotated view using `glimpse()`.
  
  To get more of a spreadsheet view, you can use the `View()` function (NB: the name starts with a *capital*, not lowercase V) or just click on the variable in the Environment pane of RStudio.
  `r unhide()`

3. Now, things are going to get a bit more tricky.
