---
title: "Correlation"
output: webex::webex_default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("webex")
```

## Task 1: Correlation matrix

How many unique pairwise correlations are there between 12 different measurements?

`r hide()`

```{r cmatrix}
choose(12, 2)
```

`r unhide()`

## Task 2: Covariance matrices

A 3x3 correlation matrix for variables $X$, $Y$, and $Z$ has the following structure (render to HTML to view):

$$
\begin{pmatrix}
{\sigma_x}^2 & \rho_{xy}\sigma_x\sigma_y & \rho_{xz}\sigma_x\sigma_z \\
\rho_{yx}\sigma_y\sigma_x & {\sigma_y}^2 & \rho_{yz}\sigma_y\sigma_z \\
\rho_{zx}\sigma_z\sigma_x & \rho_{zy}\sigma_z\sigma_y & {\sigma_z}^2 \\
\end{pmatrix}
$$

Now consider this 3x3 matrix with filled in values, and answer the following questions.

$$
\begin{pmatrix}
 4    & -4.31 &  4.45 \\
-4.31 &  9    & -6.36 \\
 4.45 & -6.36 & 16    \\
\end{pmatrix}
$$

### Task 2.1

What is the standard deviation of the $Y$ variable?

`r hide()`

```{r cvm-1}
sd_y <- 3
```

`r unhide()`

### Task 2.2

What is the covariance of $Y$ and $Z$?

`r hide()`

```{r cvm-2}
cov_yz <- -6.36
```

`r unhide()`

### Task 2.3

Is the correlation between $X$ and $Y$ negative or positive? Answer TRUE if negative, FALSE if positive.

`r hide()`

```{r cvm-3}
is_xy_negative <- TRUE
```

`r unhide()`

### Task 2.4

What is the correlation between $X$ and $Z$?

`r hide()`

```{r cvm-4}
# the sd_X = 2, sd_Z = 4 
rho_xz <- 4.45 / (sqrt(4) * sqrt(16))

rho_xz
```

`r unhide()`

## Task 3: Covariance matrices and data simulation

### Task 3.1: Construct a covariance matrix

You read a study that reports a correlation of between X and Y of $r = -.2$, and the following descriptive statistics: mean of $X = 3.3$ ($SD = 2.4$), mean of $Y = 77.1$ ($SD = 5.9$).  Adapt the code below to construct a covariance matrix that you could potentially use to simulate data from this study.

`r hide()`

```{r cvm-construct}
## my_cvmx <- matrix(0, nrow = 0, ncol = 0)
my_cvmx <- matrix(c(2.4^2, -.2 * 2.4 * 5.9,
                    -.2 * 2.4 * 5.9, 5.9^2), nrow = 2)

my_cvmx
```

`r unhide()`

### Task 3.2: Simulate data

Given the covariance matrix below and descriptive statistics below, simulate 10 observations.

$$
\begin{pmatrix}
6.25 & 1.70     \\
1.70 & 2.89 \\
\end{pmatrix}
$$

$$\bar{X} = 60$$

$$\bar{Y} = -40$$

`r hide()`

```{r}
my_Sigma <- matrix(c(6.25, 1.70, 1.70, 2.89), nrow = 2)

## simulated_data should be a 10x2 matrix
simulated_data <- MASS::mvrnorm(10, c(X = 60, Y = -40), my_Sigma)

simulated_data
```

`r unhide()`

## Task 4

Return to the "big 5" personality data that we wrangled [in the lesson on tidy data](https://dalejbarr.github.io/HU-stats-2021/02_more_wrangling/morning_tidydata/index.html#org28003ad){target="_blank"}.

Calculate the correlation matrix (the resulting matrix should be 5x5, one row/column for each of the 5 factors.) Use `correlate()` from the `corrr` package. See `vignette("using-corrr", "corrr")` for details.

`r hide()`

```{r big5, message=FALSE}
library("corrr")
library("tidyverse")

## copy the wrangling that we did to get a wide table
ocean <- read_csv("personality.csv") %>%
  select(-date) %>%
  pivot_longer(Op1:Ex9, names_to = "item", values_to = "score") %>%
  separate(item, c("subscale", "it"), 2L) %>%
  group_by(user_id, subscale) %>%
  summarize(mscore = mean(score, na.rm = TRUE), .groups = "drop")

ocean %>%
  pivot_wider(names_from = subscale, values_from = mscore) %>%
  select(-user_id) %>%
  correlate() %>%
  shave(upper = FALSE) %>%
  fashion() %>%
  knitr::kable() # make it look pretty
```

`r unhide()`